Fitting parametric univariate distributions to non censored or censored data using the R \texttt{fitdistrplus} package

### 2011-07-29 inst/doc/intro2fitdistrplus.Rnw
last non ASCII character eliminated
add of a draft of the new vignette

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%   Copyright (c) 2009 Marie Laure Delignette-Muller, Regis Pouillot, Jean-Baptiste Denis,
%   Christophe Dutang                        
%                                                                                                                                                       
%    This program is free software; you can redistribute it and/or modify                               
%    it under the terms of the GNU General Public License as published by                        
%    the Free Software Foundation; either version 2 of the License, or                                 
%    (at your option) any later version.                                                                                        
%                                                                                                                                                      
%    This program is distributed in the hope that it will be useful,                                          
%    but WITHOUT ANY WARRANTY; without even the implied warranty of                        
%    MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the                   
%    GNU General Public License for more details.                                                                    
%                                                                                                                                                         
%    You should have received a copy of the GNU General Public License                           
%    along with this program; if not, write to the                                                                            
%    Free Software Foundation, Inc.,                                                                                             
%    59 Temple Place, Suite 330, Boston, MA 02111-1307, USA                                            
%                                                                                                                                                         
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Fit parametric distributions on non-censored or censored data with fitdistrplus
%%%
%%%         Sweave vignette file
%%%

@@

\usepackage[a4paper,textwidth=18cm,textheight=27cm]{geometry}

\title{Use of the package \texttt{fitdistrplus} to specify a distribution from 
non-censored or censored data}
\author{Marie Laure Delignette-Muller, Régis Pouillot , Jean-Baptiste Denis and Christophe Dutang}

@@

Here you will find some easy examples of use of the functions of the package \texttt{fitdistrplus}. 
The aim is to show you by examples how to use these functions to help you to specify a parametric distribution 
from data corresponding to a random sample drawn from a theoretical distribution that you want to describe. 
For details, see the documentation of each function, using the R help command (ex.: \texttt{?fitdist}). 
Do not forget to load the package using the function \texttt{library} or \texttt{require}
before testing following examples.
<<>>=
library(fitdistrplus)
@

\tableofcontents

\pagebreak

\section{Specification of a distribution from non-censored continuous data}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Graphical display of the observed distribution}
%%%%%%%%%%

 First of all, the observed distribution may be plotted using the function \texttt{plotdist}.

<<plotdistx1,fig=TRUE>>=
x1<-c(6.4,13.3,4.1,1.3,14.1,10.6,
9.9,9.6,15.3,22.1,13.4,13.2,
8.4,6.3,8.9,5.2,10.9,14.4)
plotdist(x1)
@

\subsection{Characterization of the observed distribution}
%%%%%%%%%%

Descriptive parameters of the empirical distribution may be computed using the function \texttt{descdist}.
This function will also provide by default a skewness-kurtosis plot which may help you to select 
which distribution(s) to fit among the potential candidates.

<<descdistx1,fig=TRUE>>=
descdist(x1)
@

Skewness and kurtosis are known not to be robust. In order to try to
take into account the uncertainty on the estimated values of kurtosis and skewness, 
the data set may be boostrapped by fixing the argument \texttt{boot} to an integer above 10 in \texttt{descdist}. 
\texttt{boot} values of 
skewness and kurtosis corresponding to the boot nonparametric bootstrap samples are then computed and reported
on the skewness-kurtosis plot.

<<descdistx1boot,fig=TRUE>>=
descdist(x1,boot=1000)
@

\subsection{Fitting of a distribution}
%%%%%%%%%%

One or more parametric distributions may then be fitted to the data set, one at a time, using the
fonction \texttt{fitdist}. This function uses the maximum likelihood
method by default or other estimation methods described in part 5 if specified in the argument \texttt{method}.
When fitting continuous distributions, 
Cramer-von Mises, Kolmogorov-Smirnov and Anderson-Darling statistics may be computed using the function \texttt{gofstat}. 
Four goodness of fit plots are also provided.

Below is the result of the fit of a logistic distribution by maximum likelihood to the previous dataset.
<<fitdistx1l,fig=TRUE>>=
f1l<-fitdist(x1,"logis")
summary(f1l)
gofstat(f1l)
plot(f1l)
@

In that case Cramer-von Mises and Anderson-Darling tests may be used to test the adequation of the distribution to data. 
<<>>=
gofstat(f1l,print.test=TRUE)
@

Generally goodness-of-fit tests may be used carefully.
In some cases, especially on very big datasets, even if the null hypothesis is rejected a fitted distribution may be 
chosen as the best one among simple distributions to describe an empirical distribution, if the goodness-of-fit
plots do not show strong differences between empirical and theoretical distributions.

In that case, the fit seems correct, but it is easy to compare goodness-of-fit statistics
for other distributions fitted on the same dataset in order to check that other distributions could not give
a better fit. Below are computed the goodness-of-fit
statistics for logistic, lognormal, gamma, normal and weibull distributions. 
<<>>=
gofstat(fitdist(x1,"logis"))
gofstat(fitdist(x1,"lnorm"))
gofstat(fitdist(x1,"gamma"))
gofstat(fitdist(x1,"norm"))
gofstat(fitdist(x1,"weibull"))
@

Regarding the Cramer-von Mises and Kolmogorov-Smirnov statistics, 
the fit of a normal distribution seems better while regarding the 
Anderson-Darling statistic the fit of a logistic distribution seems better.
It is recommended to base the comparison of fits on the Anderson-Darling statistic when the modelling of the tails of 
a distribution is important, as it is often the case in risk assessment.

In order to choose between both distributions, the goodness-of-fit plot and the summary
of the fit of a normal chould be compared to the 
ones already obtained for the logistic distribution, but in that case the differences seem very small.
<<fitdistx1n,fig=TRUE>>=
f1n <- fitdist(x1,"norm")
plot(f1n)
summary(f1n)
plot(f1n)
@

For some distributions (see the help of \texttt{fitdist} for details), it is necessary to specify initial 
values for the distribution parameters in the argument \texttt{start} when using the maximum likelihood method.
\texttt{start} must be a named list
of parameters initial values. The names of
the parameters in \texttt{start} must correspond exactly to their definition in R or to their 
definition in a previous R code.
 The function \texttt{plotdist} may help to find correct initial values for 
the distribution parameters in non trivial cases, by an manual iterative use if necessary.

For example, below is the definition of the Gumbel distribution (also named extreme value distribution) and 
a first plot of the data set with the Gumbel distribution with arbitrary values for parameters.
<<plotdistx1gu1,fig=TRUE>>=
dgumbel<-function(x,a,b) 1/b*exp((a-x)/b)*exp(-exp((a-x)/b))
pgumbel<-function(q,a,b) exp(-exp((a-q)/b))
qgumbel<-function(p,a,b) a-b*log(-log(p))
plotdist(x1,"gumbel",para=list(a=3,b=2))
@

The same data set may be plotted with a Gumbel distribution with modified values for parameters.
<<plotdistx1gu2,fig=TRUE>>=
plotdist(x1,"gumbel",para=list(a=10,b=5))
@

And a Gumbel distribution may be fitted to data with these values for initial parameter values.
<<fitdistx1gu,fig=TRUE>>=
fgu<-fitdist(x1,"gumbel",start=list(a=10,b=5))
plot(fgu)
summary(fgu)
gofstat(fgu)
@

\subsection{Simulation of the uncertainty by boostrap}
%%%%%%%%%%

The uncertainty in the parameters of the fitted distribution may be simulated by parametric or nonparametric
boostrap using the function \texttt{boodist}. This function returns the boostrapped values of parameters which
may be plotted to visualize the bootstrap region. It also calculates the 95 percent confidence intervals for each 
parameter from the 2.5 and 97.5 percentiles
of the boostrap values of each parameter (see the help of the function \texttt{bootdist} for details).

Below is an example of the use of this function with the previous fit of the logistic distribution. 
<<bootdistf1l,fig=TRUE>>=
b1l<-bootdist(f1l)
plot(b1l)
summary(b1l)
@

\section{Specification of a distribution from non-censored discrete data}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

 A discrete data set may be considered as a continuous one for example for a large data set
 from a binomial distribution converging to a normal one. A discrete plot of the distribution 
 may also be provided, fixing the argument discrete of the function \texttt{plotdist} to \texttt{TRUE}.

<<plotdistx2,fig=TRUE>>=
x2<-rnbinom(n=100,size=2,prob=0.3)
plotdist(x2,discrete=TRUE)
@

As for continuous distributions, descriptive parameters of the empirical distribution 
may be computed using the function \texttt{descdist} which also provides a skewness-kurtosis plot 
which may help you to choose which distribution(s) to fit.

<<descdistx2,fig=TRUE>>=
descdist(x2,discrete=T,boot=1000)
@

As for continuous distributions, one or more parametric distributions may then be fitted 
to the data set by maximum likelihood or alternative methods described in part 5.

Below is the result of the fit of a Poisson distribution with the bootstrap simulations.
<<fitdistx2p,fig=TRUE>>=
f2p<-fitdist(x2,"pois")
plot(f2p)
summary(f2p)
gofstat(f2p,print.test=TRUE)
b2p<-bootdist(f2p)
summary(b2p)
@

Below is the result of the fit of a negative binomial distribution with the boostrap simulations.
<<fitdistx2nb,fig=TRUE>>=
f2n<-fitdist(x2,"nbinom")
plot(f2n)
summary(f2n)
gofstat(f2n,print.test=TRUE)
b2n<-bootdist(f2n)
summary(b2n)
@

@@

\section{Specification of a distribution from censored data}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\subsection{Graphical display of the observed distribution}
%%%%%%%%%%

First of all, the observed distribution may be plotted using the function \texttt{plotdistcens}.
Data are reported directly as segments for interval, left and right censored data, 
and as points for non-censored data. For more details, see the help of the function \texttt{plotdistcens}.

<<plotdistcensd1,fig=TRUE>>=
d1<-data.frame(
left=c(1.73,1.51,0.77,1.96,1.96,-1.4,-1.4,NA,-0.11,0.55,0.41,
    2.56,NA,-0.53,0.63,-1.4,-1.4,-1.4,NA,0.13),
right=c(1.73,1.51,0.77,1.96,1.96,0,-0.7,-1.4,-0.11,0.55,0.41,
    2.56,-1.4,-0.53,0.63,0,-0.7,NA,-1.4,0.13))
plotdistcens(d1)
@

When left or right NA-values correspond to finite value (for example 0 for
left NA-values of positive data), the arguments \texttt{leftNA} (or \texttt{rightNA})
must be affected to this finite value to ensure a correct plot of left (or right) censored
observations, as in the example below.

<<plotdistcensd2,fig=TRUE>>=
d2<-data.frame(left=10^(d1$left),right=10^(d1$right))
plotdistcens(d2,leftNA=0)
@

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%   Copyright (c) 2009 Marie Laure Delignette-Muller, Christophe Dutang                        
%                                                                                                                                                       
%    This program is free software; you can redistribute it and/or modify                               
%    it under the terms of the GNU General Public License as published by                        
%    the Free Software Foundation; either version 2 of the License, or                                 
%    (at your option) any later version.                                                                                        
%                                                                                                                                                      
%    This program is distributed in the hope that it will be useful,                                          
%    but WITHOUT ANY WARRANTY; without even the implied warranty of                        
%    MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the                   
%    GNU General Public License for more details.                                                                    
%                                                                                                                                                         
%    You should have received a copy of the GNU General Public License                           
%    along with this program; if not, write to the                                                                            
%    Free Software Foundation, Inc.,                                                                                             
%    59 Temple Place, Suite 330, Boston, MA 02111-1307, USA                                            
%                                                                                                                                                         
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Fit parametric distributions on non-censored or censored data with fitdistrplus
%%%
%%%         Sweave vignette file
%%%

@@

\usepackage[a4paper,textwidth=18cm,textheight=27cm]{geometry}
\usepackage{url}

\title{Fitting parametric univariate distributions  to non censored or
censored data using the R \texttt{fitdistrplus} package}
\author{Marie Laure Delignette-Muller and Christophe Dutang}

@@

% A enlever dans version JSS !!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
\tableofcontents

%\pagebreak

\section{Introduction}
\label{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Overview}
\label{Overview}
%%%%%%%%%%%

Fitting distributions to data is a very common task in statistics. It consists of choosing
a probability distribution that gives a good representation of a statistical
variable. It requires judgment and expertise and generally needs an iterative process of
distribution choice, parameter estimation, and quality of fit evaluation. 
Function \texttt{fitdistr} in the R package \texttt{MASS} \cite{MASS} 
is a well known general-purpose maximum-likelihood
fitting routine for the parameter estimation step in R. Other steps of
the process may be developed using R \cite{Ricci05}. Our first objective while developping
package \texttt{fitdistrplus} \cite{fitdistrplus} 
was to provide R users a set of functions dedicated to help the overall
process of fitting a univariate parametric distribution to data.

Function \texttt{fitdistr} estimates distribution parameters by maximizing the log-likelihood
using function \texttt{optim}. In some cases, other estimation methods could be prefered, 
such as maximum goodness-of-fit estimation also commonly called minimum distance estimation, and
proposed in package \texttt{actuar} 
%\cite{actuar}
with three different goodness-of-fit distances. While developping
package \texttt{fitdistrplus}, our second objective  was to extend function \texttt{fitdistr} 
by providing various
estimation methods to fit distributions in addition to maximum likelihood. Functions
were developped to enable matching moment estimation, matching quantile estimation, and 
maximum goodness-of-fit estimation (or minimum distance estimation) using eight different distances.
Moreover, package \texttt{fitdistrplus} offers the possibility to specify a user-supplied function
for optimization, 
useful in cases where optimization techniques not included in function \texttt{optim} may be more adequate.

In applied statistics, it is not uncommon to have to fit distributions to censored data. 
Function \texttt{fitdistr} 
does not enable maximum likelihood estimation from this type data. Some packages 
deal with censored data, especially survival data \cite{survival},
but those packages generally focused on specific models, enabling the fit of only one distribution or a 
restricted family of distributions. Our third objective was thus to provide R users a function 
to estimate univariate distribution parameters from censored data, whatever the type of censoring.

% A changer en cas de mise à jour de fitdistrplus !!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
This paper reviews the various features of version 0.3-0 of \texttt{fitdistrplus}. The package
is available from the Comprehensive R Archive Network at \url{http://cran.r-project.org/package=fitdistrplus}.
The development version of the package is located at R-forge as one the packages of the 
project ``Risk Assessment with R''
(\url{http://r-forge.r-project.org/projects/riskassessment/})
The following command will load the package.
<<results=hide>>=
library(fitdistrplus)
@

\subsection{Running examples}
\label{Examples}
%%%%%%%%%%%
For illustrating the use of various functions of package \texttt{fitdistrplus}, we will use
four examples published in various biological areas, 
corresponding to data sets included in the package.
% voir si on en met 3 ou 4 au final !!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!

The two first data sets correspond to the observation of a continuous variable on a random sample
of a population of interest.

The ``ground beef'' data set contains values of
serving sizes in grams, collected in a French survey, for ground beef patties consumed by 
children under 5 years old. This data set was used in a quantitative risk assessment published
in a food microbiology journal (\cite{Delignette08}).
<<echo=TRUE>>=
data(groundbeef)
str(groundbeef)
@

The ``endosulfan'' data set contains
acute toxicity values for the organochlorine pesticide endosulfan
(geometric mean of LC50 ou EC50 values in $\mu g.L^{-1}$),
tested on Australian and non-Australian laboratory-species 
(arthropods, fish or nonarthropod invertebrates) (\cite{Hose04}).
<<echo=TRUE>>=
data(endosulfan)
str(endosulfan)
@

The third data set corresponds to the observation of a discrete variable, 
the number of \emph{Toxocara cati} parasites present in digestive tract,
on a random sample of feral 
cats living on Kerguelen island (\cite{Fromont01}).
<<echo=TRUE>>=
data(toxocara)
str(toxocara)
@

The last data set corresponds to the observation of a continuous censored variable, 
the \emph{Listeria monocytogenes} microbial concentration, on a random
sample of smoked fish distributed on the Belgian market in the period
2005 to 2007 (\cite{Busschaert10}).
Censored data are coded within 2 columns named left and right, describing
each observed value of \emph{Listeria monocytogenes} concentration 
(in $CFU.g^{-1}$) as an interval. 
The left column contains either \texttt{NA} for
left censored observations, the left bound of the interval for interval censored
observations, or the observed value for non-censored observations. The right
column contains either \texttt{NA} for right censored observations, the right bound of
the interval for interval censored observations, or the observed value for noncensored
observations.
<<echo=TRUE>>=
data(smokedfish)
str(smokedfish)
@

\section{Choice of candidate distributions}
\label{Choice}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Before fitting one or more distributions to a data set, it is generally
necessary to choose good candidates among a predefined family of distributions.
To help the user in this preliminary task, we developed functions to plot and
characterise empirical distributions.

\subsection{Graphical display of the observed distribution}
\label{Graphical}
%%%%%%%%%%

First of all, an empirical distribution may be plotted using classical R function
or using Function \texttt{plotdist} which provides plots
in density and in cdf as done in (Figure~\ref{plotdistcont}) for a continuous variable:

<<echo=TRUE, fig=FALSE>>=
plotdist(groundbeef$serving)
@
\begin{figure}[htb]
  \centering
<<echo=FALSE, fig=TRUE>>=
plotdist(groundbeef$serving)
@
  \caption{Density and cdf plots of an empirical distribution for a continuous variable}
  \label{plotdistcont}
\end{figure}

In some cases a discrete variable may be plotted as a continuous one, for example for a large data set
from a binomial distribution converging to a normal one, but
Function \texttt{plotdist} also proposes specific plots in density and in cdf
for discrete variables (Figure~\ref{plotdistdisc}):

<<echo=TRUE, fig=FALSE>>=
plotdist(toxocara$number,discrete = TRUE)
@
\begin{figure}[htb]
  \centering
<<echo=FALSE, fig=TRUE>>=
plotdist(toxocara$number,discrete = TRUE)
@
  \caption{Density and cdf plots of an empirical distribution for a discrete variable}
  \label{plotdistcont}
\end{figure}

%\clearpage

\subsection{Empirical basis for selecting candidate distributions}
\label{Empirical}
%%%%%%%%%%
Function \texttt{descdist} 
Descriptives statistics may help the choice of good candidates to describe an empirical distribution
among a family of parametric distributions. Especially the skewness and kurtosis are useful for this
purpose. The concept of skewness relates to deviations from symmetry of the distribution.
The normal distribution has a skewness of zero. A positive (resp. negative) skewness indicates 
that the right (resp. left) tail of the distribution is more extended than the left (resp. right) one.
The concept of kurtosis relates to the tail weight. The normal distribution has a kurtosis of 3.
Distributions with a higher kurtosis are said to be leptokurtic, with heavier tails, such as the 
logistic distribution, while distributions with a smaller kurtosis are said platykurtic, with lighter 
tails, such as the uniform distribution.

Function \texttt{descdist} provides calculations of classical descriptive statistics
(minimum, maximum, median, mean, sample sd) and by default 
unbiased estimations of skewness and 
Pearsons's kurtosis values. Nevertheless estimations of skewness and kurtosis
are unbiased only for normal distributions and estimated values are thus only indicative.
A skewness-kurtosis plot such as the one proposed by \cite{Cullen99} is also provided for the 
empirical distribution (Figure~\ref{Cullenplotcont}). 
On this plot, values for common distributions are displayed as a tools 
to help the choice of distributions to fit to data. For some distributions (normal, uniform,
logistic, exponential for example), there is only one possible value for the skewness and the kurtosis
 and the distribution is thus represented by a point on the plot. For other distributions, 
areas of possible values are represented, consisting in lines (as for gamma and lognormal distributions), 
or larger areas (as for beta distribution).

Skewness and kurtosis are known not to be robust. In order to take into account the uncertainty 
of the estimated values of kurtosis and skewness from data, the data set may be boostraped by 
fixing the argument \texttt{boot} to an integer above 10. Values of skewness and kurtosis 
corresponding to bootstrap samples are then computed and reported on the 
skewness-kurtosis plot.

<<echo=TRUE, fig=FALSE>>=
descdist(groundbeef$serving,boot=1000)
@
\begin{figure}[htb]
  \centering
<<echo=FALSE, results=hide, fig=TRUE>>=
descdist(groundbeef$serving,boot=1000)
@
  \caption{Skewness-kurtosis plot for a continuous variable}
  \label{Cullenplotcont}
\end{figure}

For discrete variable, skewness and kurtosis values or set of values of Poisson and negative binomial
distributions are represented in the skewness-kurtosis plot (Figure~\ref{Cullenplotdisc}), 
together with values for the
normal distribution, to which discrete distributions may converge.

<<echo=TRUE, fig=FALSE>>=
descdist(toxocara$number,discrete = TRUE,boot=1000)
@
\begin{figure}[htb]
  \centering
<<echo=FALSE, results=hide, fig=TRUE>>=
descdist(toxocara$number,discrete = TRUE,boot=1000)
@
  \caption{Skewness-kurtosis plot for a continuous variable}
  \label{Cullenplotdisc}
\end{figure}

\clearpage

\section{Fit of a distribution}
\label{FIT}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Parameter estimation}
\label{Estim}
%%%%%%%%%%
Once selected, one or more parametric distributions may be fitted to the data set, one at a time, using
Function \texttt{fitdist}. By default, distribution parameters $\theta$ are estimated by maximizing the
likelihood defined as:
  \begin{equation}
    \label{likelihood}
    L(\theta)=\prod_{i=1}^n f(y_{i}|\theta)
  \end{equation}
with $y_{i}$ the $n$ observations of variable $y$ and $f$ the density function of the fitted 
parametric distribution.

The other proposed estimation methods are described in Section~\ref{Alternatives}.

Function \texttt{fitdist} returns the results of the fit of any parametric distribution 
to a data set as an S3 class object that may be easily printed, 
summarized or plotted (see Figure~\ref{fitgamma} in Section~\ref{GOF plots}). 
The parametric distribution must be a classically defined R distributions,
with at least d, p and q functions respectively for the density cdf and quantile functions
(for example \texttt{dnorm}, \texttt{pnorm} and \texttt{qnorm} for the normal distribution).
The name of the fitted distribution is specified in the first argument by its classical 
abbreviation used as the second part of d, p and q functions (for example ``norm''
for the normal distribution. Numerical results returned by Function \texttt{fitdist} are 
parameter estimates with estimated standard errors computed from the estimate
of the Hessian matrix at the maximum likelihood solution, correlation matrix
between parameter estimates, the loglikelihood, the Akaïke and the Schwarz information criteria 
(so called AIC and BIC).
<<echo = TRUE,fig = FALSE>>=
fw <- fitdist(groundbeef$serving,"weibull")
print(fw)
summary(fw)
@

The same procedure is required to fit a discrete distribution. As an example, 
using ``toxocara'' data set, Poisson and
negative distributions may be easily fitted and AIC values compared, in this case
giving the preference
to the negative binomial distribution, with a much smaller AIC value.
<<echo = TRUE,fig = FALSE>>=
(fp <- fitdist(toxocara$number,"pois"))
(fnb <- fitdist(toxocara$number,"nbinom"))
fp$aic
fnb$aic
@

For some distributions (see the help of \texttt{fitdist} for details), it is necessary to specify initial 
values for the distribution parameters in the argument \texttt{start} when using the maximum likelihood method.
\texttt{start} must be a named list
of parameters initial values. The names of
the parameters in \texttt{start} must correspond exactly to their definition in R or in a user-supplied R code.
Function \texttt{plotdist} (see Section~\ref{GOF plots}), 
which can plot any parametric distribution with specified parameter values
in argument \texttt{para} may help to find correct initial values for 
the distribution parameters in non trivial cases, by iterative calls if necessary (see~\cite{fitdistrplus}
for examples).

% Faudra trouver un autre exemple ou une autre distribution ou enlever cette partie

%As an example, here is the fit of a user-supplied distribution, the Gumbel distribution 
%(also named extreme value distribution) to an arbitrary data set, preceded by 
%a plot of this data set with the Gumbel distribution which parameters 
%were fixed at initial values manually obtained after iterative plots.

%<<echo=TRUE,fig=TRUE>>=
%x1 <- c(6.4, 13.3, 4.1, 1.3, 14.1, 10.6, 9.9, 9.6, 15.3, 22.1,
%13.4, 13.2, 8.4, 6.3, 8.9, 5.2, 10.9, 14.4)
%dgumbel<-function(x,a,b) 1/b*exp((a-x)/b)*exp(-exp((a-x)/b))
%pgumbel<-function(q,a,b) exp(-exp((a-q)/b))
%qgumbel<-function(p,a,b) a-b*log(-log(p))
%plotdist(x1,"gumbel",para=list(a=10,b=5))
%fitdist(x1,"gumbel",start=list(a=10,b=5))
%@
%\begin{figure}[htb]
%  \centering
%<<echo=FALSE, results=hide, fig=TRUE>>=
%plotdist(x1,"gumbel",para=list(a=10,b=5))
%@
%  \caption{Preliminar plot of a distribution against data to find initial values for parameters}
%  \label{plotdist}
%\end{figure}

\subsection{Goodness-of-fit plots}
\label{GOF plots}
%%%%%%%%%%
The plot of an object of class \texttt{fitdist} provides two types of results depending of the nature 
of the distribution, continuous or discrete. For continuous distributions, 
four goodness-of-fit plots are provided : a draw of pdf curve and histogram together, an cdf plot
of both empirical and theoretical distributions, a Q-Q plot
(plot of the quantiles of the theoretical fitted distribution (x-axis) against the empirical quantiles of the data)
and a P-P plot (i.e. for each value of the data set, plot of the cumulative density function of the fitted distribution
(x-axis) against the empirical cumulative density function (y-axis)) are also given (\cite{Cullen99}).
As an exemple, let us look at the plot of the previous fit of a weibull distribution to ``groundbeef'' data set.

<<echo=TRUE, fig=FALSE>>=
plot(fw)
@
\begin{figure}[htb]
  \centering
<<echo=FALSE, results=hide, fig=TRUE>>=
plot(fw)
@
  \caption{Plot of the fit of a continuous distribution}
  \label{plotcontfit}
\end{figure}

For continuous distributions, Function \texttt{cdfcomp} enables the visual comparison
of empirical and theoretical cumulative distributions for various distributions fitted
on a same data set. Function \texttt{cdfcomp} must be called with a first argument 
corresponding to a list of objects of class \texttt{fitdist}, and optionnaly
further arguments to customize the plot, as in the following example comparing the fit of
Weibull, lognormal and gamma distributions to ``groundbeef'' data set.
<<echo=TRUE, fig=FALSE>>=
    fg <- fitdist(groundbeef$serving,"gamma")
    fln <- fitdist(groundbeef$serving,"lnorm")
    cdfcomp(list(fw,fln,fg),legendtext=c("Weibull","lognormal","gamma"),
    xlab="serving sizes (g)",lwd=2)
@
\begin{figure}[htb]
  \centering
<<echo=FALSE, results=hide, fig=TRUE>>=
    cdfcomp(list(fw,fln,fg),legendtext=c("Weibull","lognormal","gamma"),
    xlab="serving sizes (g)",lwd=2)
@
  \caption{Comparison of CDF plots of various distributions fitted on continuous data}
  \label{compcdf}
\end{figure}

In such a plot, data may be represented in a log scale when required, by just fixing the argument
\texttt{xlogscale} to \texttt{TRUE} in the call to \texttt{cdfcomp}.
% Ajouter ex endosulfan avec la bibliothèque actuar !!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
% mais comment faire dans la vignette : indiquer que fitdistrplus nécessite actuar????????????????????????????

For discrete distributions, the plot of an object of class \texttt{fitdist} provides
two goodness-of-fit plots comparing empirical and theoretical distributions
in pdf and in cdf.
As an exemple, let us look at the plot of the previous fit of a negative binomial distribution 
to ``toxocara'' data set.

<<echo=TRUE, fig=FALSE>>=
plot(fnb)
@
\begin{figure}[htb]
  \centering
<<echo=FALSE, results=hide, fig=TRUE>>=
plot(fnb)
@
  \caption{Plot of the fit of a discrete distribution}
  \label{plotdiscfit}
\end{figure}

\subsection{Measures of goodness-of-fit}
\label{GOF measures}
%%%%%%%%%%
When fitting continuous distributions, 
Cramer-von Mises, Kolmogorov-Smirnov and Anderson-Darling statistics may be computed using the function \texttt{gofstat}
as defined by Stephens (\cite{Stephens86}).
<<echo=TRUE,fig=FALSE>>=
gofstat(fw)
@

% ADD the definition of these statistics !!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
EQUATIONS TO BE ADDED

As giving  more weight to distribution tails, 
Anderson-Darling statistics is of special interest where it is important to place equal 
emphasis on fitting a distribution at the tails as well as the main body, 
as it is often the case in risk assessment \cite{Cullen99,Vose10}.

% ADD AN EXAMPLE with comparison !!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
Nevertheless, this statistics should 
be used cautiously when comparing fits of various distributions, keeping in mind that the weighting of each
cdf quadratic difference is dependent of the theoretical distribution.

When fitting discrete distributions, the Chi-squared statistic is computed by 
Function \texttt{gofstat} using cells defined 
by the argument \texttt{chisqbreaks} or cells automatically defined from the data in order 
to reach roughly the same number of observations per cell, roughly equal to the argument
\texttt{meancount}, or sligthly more if there are some ties. 
The choice to define cells from the empirical distribution (data) and not from the
theoretical distribution was done to enable the comparison of Chi-squared values obtained
with different distributions fitted on a same dataset.
If arguments \texttt{chisqbreaks} and \texttt{meancount} 
are both omitted, \texttt{meancount} is fixed in order to obtain roughly $(4n)^{2/5}$ cells, 
with $n$ the length of the dataset~\cite{Vose10}. Using this default option with 
the fit of a negative binomial distribution to ``toxocara'' data set gives following results :
<<echo=TRUE,fig=FALSE>>=
gofstat(fnb)
@

Among its returned values, Function \texttt{gofstat} provides a table with observed and theoretical 
counts used for the Chi-squared calculations:
<<echo=TRUE,fig=FALSE>>=
gofstat(fnb)$chisqtable
@

Even if specifically recommended for discrete distributions, the Chi-squares statistic may also be used for
continuous distributions (see~\cite{fitdistrplus} for examples).

\subsection{Goodness-of-fit tests}
\label{GOF tests}
%%%%%%%%%%

For continuous distributions, an approximate Kolmogorov-Smirnov test is 
performed by assuming the distribution parameters known. The critical value defined by Stephens~\cite{Stephens86} 
for a completely specified distribution is used to reject or not the 
distribution at the significance level 0.05. Because of this approximation, the result of the test
(decision of rejection of the distribution or not) is returned only for datasets with more 
than 30 observations. Note that this approximate test may be too conservative.

For datasets with more than 5 observations and for continuous distributions for 
which the test is described by Stephens~\cite{Stephens86} (normal, lognormal,
exponential, Cauchy, gamma, logistic and Weibull),
the Cramer-von Mises and Anderson-darling tests are performed as described by Stephens~\cite{Stephens86}. 
Those tests take into 
account the fact that the parameters are not known but estimated from the data. The result is the 
decision to reject or not the distribution at the significance level 0.05. Both tests are available
only for maximum likelihood estimations.

When the Chi-squared statistic is computed (for discrete or optionnaly continuous distributions), 
and if the degree of freedom (nb of cells - nb of parameters - 1)  of the corresponding distribution 
is strictly positive, the p-value of the Chi-squared test is returned.

Goodness-of-fit tests may be used carefully. As for any null-hypothesis significance
test, the non reject of the null hypothesis dose not imply its acceptation. However, this
misinterpretation of p-values is very common and comes from the wrong assumption that
absence of evidence is evidence of absence~\cite{Altman96}.
On the contrary, in some cases, especially on very big datasets, 
even if the null hypothesis is rejected a fitted distribution may be 
chosen as the best one among simple distributions to describe an empirical distribution, if the goodness-of-fit
plots do not show strong differences between empirical and theoretical distributions.

% ADD an example to argument that !!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!

Now let us look at the Chi-squared test results for the fit of a negative binomial distribution 
to ``toxocara'' data set :
<<echo=TRUE,fig=FALSE>>=
gofstat(fnb,print.test = TRUE)
@

A warning message appears as one of the theoretical counts is under 5 using the default breaks 
(see Section~\ref{GOF measures}). In order to solve this problem, one may specify breaks 
more adapted for the realization of the test.
<<echo=TRUE,fig=FALSE>>=
gofstat(fnb,chisqbreaks=c(0,1,4,8,20),print.test=TRUE)$chisqtable
@

@@

%\clearpage

\section{The special case of censored data}
\label{censored}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\subsection{Graphical display of the observed distribution}
\label{censored graph}
%%%%%%%%%%

Using censored data such as those coded in the ``smokedfish'' data set,
the empirical distribution may be plotted using the function \texttt{plotdistcens}.
Data are reported directly as segments for interval, left and right censored data, 
and as points for non-censored data. Before plotting, observations are ordered and a rank r
is associated to each of them. Left censored observations are ordered
first, by their right bounds. Interval censored and non censored observations
are then ordered by their mid-points and, at last, right censored observations are
ordered by their left bounds. If argument \texttt{leftNA} (resp. \texttt{rightNA}) is finite,
left censored (resp. right censored) observations are considered as interval censored
observations and ordered by mid-points with non-censored and interval censored data.
It is sometimes necessary to fix \texttt{leftNA} or \texttt{rightNA} to a realistic 
extreme value, even if not exactly known, to obtain a reasonable global ranking of 
observations. After ranking, each of the n observations is plotted as a point (one x-value) 
or a segment (an interval of possible x-values),
with an y-value equal to r/n, r being the rank of each observation in the global ordering
previously described.

Let us see the sulting plot for ``smokedfish'' data set after classical transformation
of microbila counts in decimal logarithm.
